---
title: "Getting Started with R"
author: "Dr. Joshua Lambert"
format:
  html:
    toc: true
    toc-title: "Table of Contents"
    code-fold: false
    code-tools: true
    css: ../styles.css
---

[**Download R Code for this Lecture**](code/getting-started-with-r.R){ .btn .btn-primary }



# Getting Started with R

Welcome to R! This guide will help you get started with R and RStudio, even if you have no prior programming experience. R is a powerful, free statistical software that is widely used in research, including nursing and healthcare.

## Why R?

R offers several advantages for nursing research:

- **Free and Open Source**: No licensing fees, available to everyone
- **Powerful**: Can handle simple to complex statistical analyses
- **Reproducible**: Your analysis is documented in code, making it easy to replicate
- **Widely Used**: Large community of users and extensive online resources
- **Flexible**: Can create publication-quality graphics and reports

## Installing R and RStudio

You'll need to install two programs: R (the statistical software) and RStudio (a user-friendly interface for R).

### Step 1: Install R

1. Go to [https://cran.r-project.org/](https://cran.r-project.org/)
2. Click on the download link for your operating system:
   - **Windows**: Click "Download R for Windows" → "base" → Download the latest version
   - **Mac**: Click "Download R for macOS" → Download the appropriate version for your Mac
3. Run the installer and follow the default installation options

### Step 2: Install RStudio

1. Go to [https://posit.co/download/rstudio-desktop/](https://posit.co/download/rstudio-desktop/)
2. Click "Download RStudio Desktop" (the free version)
3. Run the installer and follow the default installation options

> **💡 R Tip**: Always install R *before* RStudio. RStudio needs R to be installed first to work properly.

## Understanding the RStudio Interface

When you open RStudio, you'll see four main panes:

### 1. Script Editor (Top Left)
- Where you write and save your R code
- Create a new script: File → New File → R Script
- Save your work here for future use

### 2. Console (Bottom Left)
- Where R executes commands and displays results
- You can type commands directly here, but they won't be saved
- Output and messages appear here

### 3. Environment/History (Top Right)
- **Environment**: Shows all objects (data, variables) currently in memory
- **History**: Shows all commands you've run

### 4. Files/Plots/Packages/Help (Bottom Right)
- **Files**: Browse files on your computer
- **Plots**: View graphs and visualizations
- **Packages**: Manage R packages (extensions)
- **Help**: Access documentation

> **💡 R Tip**: You can customize the pane layout in Tools → Global Options → Pane Layout.

## R Basics

### The Assignment Operator

In R, we use `<-` to assign values to objects (variables):

```{r}
# Assign the value 5 to an object called 'x'
x <- 5

# View the value of x
x

# You can also use = for assignment, but <- is preferred in R
y <- 10
y
```

> **💡 R Tip**: The keyboard shortcut for `<-` is Alt + - (Windows) or Option + - (Mac).

### Data Types

R has several basic data types:

```{r}
# Numeric (numbers)
age <- 45
weight <- 68.5

# Character (text - always in quotes)
name <- "Patient A"
diagnosis <- "Hypertension"

# Logical (TRUE or FALSE)
is_smoker <- TRUE
has_diabetes <- FALSE

# Check the type of an object
class(age)
class(name)
class(is_smoker)
```

### Vectors

Vectors are collections of values of the same type. We create them using `c()` (combine):

```{r}
# Numeric vector
ages <- c(25, 34, 45, 56, 67)
ages

# Character vector
treatments <- c("Placebo", "Drug A", "Drug B", "Placebo", "Drug A")
treatments

# Access individual elements using brackets []
ages[1]  # First element
ages[3]  # Third element

# Access multiple elements
ages[c(1, 3, 5)]  # First, third, and fifth elements
```

> **💡 R Tip**: R uses 1-based indexing, meaning the first element is at position 1 (not 0 like some other languages).

### Data Frames

Data frames are like spreadsheets - they have rows and columns. This is how we typically work with data in R:

```{r}
# Create a simple data frame
patient_data <- data.frame(
  patient_id = c(1, 2, 3, 4, 5),
  age = c(25, 34, 45, 56, 67),
  treatment = c("Placebo", "Drug A", "Drug B", "Placebo", "Drug A"),
  outcome = c(5.2, 6.8, 7.1, 5.5, 6.9)
)

# View the data frame
patient_data

# View structure of the data frame
str(patient_data)

# View first few rows
head(patient_data)

# Access a column using $
patient_data$age

# Access a column using brackets
patient_data[, "treatment"]

# Access a specific cell [row, column]
patient_data[2, 3]  # Row 2, Column 3
```

### Functions

Functions perform operations. They have a name followed by parentheses containing arguments:

```{r}
# Calculate mean (average)
mean(ages)

# Calculate standard deviation
sd(ages)

# Calculate median
median(ages)

# Summary statistics
summary(ages)

# Count number of elements
length(ages)

# Create a table of frequencies
table(treatments)
```

> **💡 R Tip**: To learn about any function, type `?function_name` in the console (e.g., `?mean`).

## Working with Data

### Setting Your Working Directory

The working directory is where R looks for files and saves output:

```{r, eval=FALSE}
# Check current working directory
getwd()

# Set working directory (use your own path)
setwd("C:/Users/YourName/Documents/R_Projects")

# Or use Session → Set Working Directory → Choose Directory in RStudio
```

> **💡 R Tip**: Use forward slashes `/` in file paths, even on Windows. Or use double backslashes `\\`.

### Reading Data from CSV Files

CSV (Comma-Separated Values) files are common for data storage. There are three main ways to read them:

#### 1. Direct File Path
Best for reproducibility (your script will always know where to look).

```{r, eval=FALSE}
# Read a CSV file by providing the path
my_data <- read.csv("data/patient_data.csv")
```

#### 2. Interactive Selection (file.choose)
Useful if you don't know the exact path or want to "browse" for the file.

```{r, eval=FALSE}
# This will open a window for you to select the file manually
my_data <- read.csv(file.choose())
```

> **💡 R Tip**: `file.choose()` is great for beginners, but for formal research, typing the path (Method 1) is better because it makes your work "reproducible" - anyone can run your script without having to click through windows.

#### 3. RStudio "Import Dataset" Tool
The easiest way for beginners. RStudio has a built-in "point-and-click" importer:

1. In the **Environment** pane (top right), click **Import Dataset**.
2. Select **From Text (base)...** (for CSV) or **From Excel...**.
3. Browse to your file.
4. Preview the data and check the options (e.g., Heading, Separator).
5. Click **Import**.

> **💡 R Tip**: When you use the Import Tool, RStudio actually runs the `read.csv()` or `read_excel()` code in the console. You should copy that code and paste it into your script so you can re-run it later!

#### Examining the Data
Once imported, always check your data:

```{r, eval=FALSE}
# View the first few rows
head(my_data)

# View the structure (variable names and types)
str(my_data)

# View the entire dataset in a spreadsheet-like viewer
View(my_data)
```

### Reading Data from Excel Files

To read Excel files, you need the `readxl` package:

```{r, eval=FALSE}
# Install the package (only needed once)
install.packages("readxl")

# Load the package (needed each session)
library(readxl)

# Read an Excel file
my_data <- read_excel("data_file.xlsx")

# Read a specific sheet
my_data <- read_excel("data_file.xlsx", sheet = "Sheet1")
```

## Basic Data Manipulation

### Subsetting Data

```{r}
# Create example data
patient_data <- data.frame(
  patient_id = 1:10,
  age = c(25, 34, 45, 56, 67, 23, 45, 56, 78, 34),
  treatment = rep(c("Placebo", "Drug A"), 5),
  outcome = c(5.2, 6.8, 7.1, 5.5, 6.9, 5.0, 7.2, 6.5, 7.8, 6.2)
)

# Select specific columns
patient_data[, c("patient_id", "age")]

# Select rows where age > 40
patient_data[patient_data$age > 40, ]

# Using subset() function (easier to read)
subset(patient_data, age > 40)

# Select specific columns for subset
subset(patient_data, age > 40, select = c(patient_id, age, outcome))
```

### Creating New Variables

```{r}
# Add a new column
patient_data$age_group <- ifelse(patient_data$age >= 50, "Older", "Younger")

# View the updated data
patient_data

# Create a new variable based on calculations
patient_data$outcome_doubled <- patient_data$outcome * 2

patient_data
```

### Basic Transformations

```{r}
# Log transformation
patient_data$log_outcome <- log(patient_data$outcome)

# Standardization (z-scores)
patient_data$outcome_z <- scale(patient_data$outcome)

# View results
head(patient_data)
```

## R Packages

Packages are collections of functions that extend R's capabilities. Think of them as apps for R.

### Installing Packages

```{r, eval=FALSE}
# Install a package (only needed once)
install.packages("ggplot2")

# Install multiple packages at once
install.packages(c("ggplot2", "dplyr", "car"))
```

### Loading Packages

```{r, eval=FALSE}
# Load a package (needed each time you start R)
library(ggplot2)

# Alternative way to load
require(ggplot2)
```

### Commonly Used Packages in This Course

- **ggplot2**: Advanced data visualization
- **car**: Companion to Applied Regression (for ANOVA/regression diagnostics)
- **ez**: Easy analysis and visualization of factorial experiments
- **pwr**: Power analysis
- **readxl**: Reading Excel files
- **tidyr**: Data reshaping and tidying

> **💡 R Tip**: You only need to install a package once, but you need to load it with `library()` every time you start a new R session.

## Getting Help

### Built-in Help

```{r, eval=FALSE}
# Get help on a function
?mean
help(mean)

# Search for help on a topic
??regression

# Get examples of how to use a function
example(mean)
```

### Understanding Error Messages

Don't panic when you see errors! They're R's way of telling you something needs fixing.

**Common errors:**

```{r, eval=FALSE}
# Error: object not found
print(my_variable)
# Fix: Make sure you've created the object and spelled it correctly

# Error: could not find function
myfunction()
# Fix: Check spelling, or load the package that contains the function

# Error: unexpected symbol
x < - 5  # Space in assignment operator
# Fix: Remove the space: x <- 5
```

### Online Resources

- **RStudio Cheat Sheets**: [https://posit.co/resources/cheatsheets/](https://posit.co/resources/cheatsheets/)
- **Stack Overflow**: Search for R questions and answers
- **R Documentation**: [https://www.rdocumentation.org/](https://www.rdocumentation.org/)
- **Quick-R**: [https://www.statmethods.net/](https://www.statmethods.net/)

## Best Practices

### 1. Comment Your Code

Use `#` to add comments explaining what your code does:

```{r}
# Calculate the mean age of patients
mean_age <- mean(patient_data$age)

# This is a comment - R ignores everything after #
```

### 2. Use Meaningful Names

```{r, eval=FALSE}
# Good
patient_age <- 45
treatment_group <- "Drug A"

# Not as good
x <- 45
tg <- "Drug A"
```

### 3. Organize Your Scripts

```{r, eval=FALSE}
# ============================================
# Project: Blood Pressure Study
# Author: Your Name
# Date: 2026-01-05
# ============================================

# Load packages
library(ggplot2)
library(car)

# Read data
bp_data <- read.csv("blood_pressure.csv")

# Data cleaning
# ... your code ...

# Analysis
# ... your code ...

# Visualization
# ... your code ...
```

### 4. Save Your Work

- Save your R scripts (File → Save) with a `.R` extension
- Save your data frames if needed:

```{r, eval=FALSE}
# Save data frame as CSV
write.csv(patient_data, "cleaned_data.csv", row.names = FALSE)

# Save R objects
save(patient_data, file = "my_data.RData")

# Load saved R objects
load("my_data.RData")
```

## Comprehensive Example: Complete Data Analysis Workflow

Let's put it all together with a complete example:

```{r}
# ============================================
# Example: Analyzing Patient Recovery Times
# ============================================

# Step 1: Create sample data (normally you'd read this from a file)
recovery_data <- data.frame(
  patient_id = 1:30,
  age = c(45, 52, 38, 61, 49, 55, 42, 58, 47, 53,
          39, 44, 56, 50, 43, 59, 48, 51, 46, 54,
          41, 57, 62, 40, 60, 37, 63, 36, 64, 35),
  treatment = rep(c("Standard", "Enhanced", "Intensive"), each = 10),
  recovery_days = c(12, 15, 10, 18, 14, 16, 11, 17, 13, 15,
                    9, 11, 8, 12, 10, 13, 9, 11, 10, 12,
                    7, 9, 6, 8, 7, 9, 6, 8, 7, 8)
)

# Step 2: Examine the data
print("First few rows:")
head(recovery_data)

print("\nData structure:")
str(recovery_data)

print("\nSummary statistics:")
summary(recovery_data)

# Step 3: Basic descriptive statistics by group
print("\nMean recovery days by treatment:")
tapply(recovery_data$recovery_days, recovery_data$treatment, mean)

print("\nStandard deviation by treatment:")
tapply(recovery_data$recovery_days, recovery_data$treatment, sd)

# Step 4: Create a new variable (age group)
recovery_data$age_group <- ifelse(recovery_data$age >= 50, "50+", "Under 50")

# Step 5: Create a simple visualization
boxplot(recovery_days ~ treatment, 
        data = recovery_data,
        main = "Recovery Days by Treatment",
        xlab = "Treatment Type",
        ylab = "Recovery Days",
        col = c("lightblue", "lightgreen", "lightcoral"))

# For more advanced visualizations, see the [Graphing Your Data](graphing-your-data.qmd) lecture.

# Step 6: Basic statistical test (we'll learn more about this later)
# ANOVA to test if recovery days differ by treatment
# Learn more in the [Introduction to ANOVA](introduction-to-anova.qmd) lecture.
anova_result <- aov(recovery_days ~ treatment, data = recovery_data)
print("\nANOVA Results:")
summary(anova_result)

# Step 7: Save the cleaned data (if this were a real analysis)
# write.csv(recovery_data, "recovery_data_cleaned.csv", row.names = FALSE)
```

## Summary

You now know how to:

✅ Install R and RStudio  
✅ Navigate the RStudio interface  
✅ Understand basic R syntax and data types  
✅ Create and manipulate vectors and data frames  
✅ Read data from CSV and Excel files  
✅ Perform basic data manipulation  
✅ Install and load packages  
✅ Get help when you need it  
✅ Follow best practices for reproducible research  

## Next Steps

Now that you're familiar with R basics, you're ready to move on to the statistical methods covered in this course. Each lecture will build on these foundations, showing you how to use R for specific analyses.

> **💡 R Tip**: Don't try to memorize everything! Keep this guide handy as a reference. The more you practice, the more natural R will become.

**Happy coding! 🎉**
