---
title: "Graphing Your Data"
author: "Dr. Joshua Lambert"
output: html
format:
  html:
    toc: true
    toc-title: "Table of Contents"
    code-fold: false
    code-tools: true 
---

# Graphing Your Data

## Identifying Differences in Types of Data and How to Graph

### Why Graph Data? Benefits

Graphs are crucial in data analysis and presentation because they help:

-   **Quick Interpretation:** Graphs provide a quick overview of data trends, making them easier to interpret than tables.

-   **Pattern Identification:** They help identify patterns, relationships, and outliers.

-   **Hypothesis Generation:** Aids in forming research hypotheses.

-   **Audience Engagement:** Captures and maintains the audience's attention.

-   **Decision Support:** Facilitates evidence-based decision-making.

### Data Types

1.  **Categorical Data:**

    -   **Nominal:** No inherent order or ranking (e.g., coffee, tea, water).\
    -   **Ordinal:** Has a meaningful order or ranking (e.g., small, medium, large).

    **Pros:** Easy to categorize and analyze.\
    **Cons:** Limited statistical techniques apply, and numerical differences are not meaningful.

2.  **Continuous Data:**

    -   **Interval:** Has equal intervals between values but no true zero (e.g., temperature, dates).\
    -   **Ratio:** Has a meaningful zero point and equal intervals (e.g., weight, strength, pressure).

    **Pros:** Wide range of statistical techniques available.\
    **Cons:** Requires more complex handling due to scale.

### Types of Graphs

1.  **Univariate Graphs:**

    -   **Definition:** Focuses on a single variable.
    -   **Examples:** Histograms, Box plots.

    **Pros:**

    -   Simple and intuitive.
    -   Provides distributional insights.

    **Cons:**

    -   Limited to understanding one variable.

2.  **Bivariate Graphs:**

    -   **Definition:** Displays the relationship between two variables.
    -   **Examples:** Scatter plots, Box plots.

    **Pros:**

    -   Useful for studying relationships and trends.

    **Cons:**

    -   Might not reveal hidden multivariate trends.

3.  **Multivariate Graphs:**

    -   **Definition:** Involves multiple variables.
    -   **Examples:** Scatter plot matrices, Mosaic plots, Treemaps.

    **Pros:**

    -   Effective for uncovering complex relationships.

    **Cons:**

    -   Interpretation can be challenging without proper labels.

## Stephen Few's Graphic Data Display Key Points

-   **Know Your Audience:** Understand how your audience processes information to tailor your visualizations effectively.
-   **Define Your Message:** Clearly convey your message simply and quickly.
-   **Use Visual Elements Appropriately:** Combine words, numbers, and images meaningfully.
-   **Principles of Excellence:** Based on Edward Tufte's principles of good graphical design:
    -   **Enforce Visual Comparisons:** Allow comparisons between data points.
    -   **Show Causality:** Highlight causal relationships if possible.
    -   **Integrate All Visual Elements:** Include relevant text and numbers to provide context.
    -   **Content-Driven Design:** Ensure design is driven by data quality, relevance, and integrity.

## Principles of Good and Bad Graphical Design

### Tips and Tricks

-   **Graphical Excellence:**
    -   **Show Multivariate Data:** Present data across multiple dimensions.
    -   **Integrate All Visual Elements:** Include text, numbers, and images cohesively.
    -   **Use Quality, Relevant, and Honest Data:** Ensure data integrity and honesty.
-   **Common Mistakes:**
    -   **Distorting Data Meaning:** Misleading visuals distort interpretation.
    -   **Incorrect Scaling:** Scaling errors can misrepresent data trends.
    -   **Poor Data-to-Ink Ratio:** Avoid excessive chart junk that doesn't add value.

[The 27 Worst Charts Of All Time](https://www.businessinsider.com/the-27-worst-charts-of-all-time-2013-6#unordered-perplexingly-circular-and-difficult-to-ascertain-any-information-from-this-globe-and-mail-chart-they-went-out-of-their-way-to-make-this-flashy-but-difficult-to-actually-read-we-get-what-they-were-going-for-but-this-chart-is-precisely-why-humanity-invented-tables-6)

<iframe width="560" height="315" src="https://www.youtube.com/embed/FcllfpC7kAY" frameborder="0" allowfullscreen>

</iframe>

## Graph Types and How to Use Them

### Histogram

```{r, echo=FALSE, message=FALSE}
  # Create a histogram in R
        data <- rnorm(100)
        hist(data, main = "Histogram of Data", xlab = "Values", col = "blue", border = "black")
```

-   **Definition:** Graphical representation of data distribution using bins of equal length to count frequencies.
-   **Pros:**
    -   Helps visualize the distribution shape.
    -   Easy to compare different distributions.
    -   Identifies outliers and skewness.
-   **Cons:**
    -   Bin size selection can lead to over or under-smoothing.
    -   Not suitable for small data sets.

```{r, eval=FALSE}
# Create a histogram in R
        data <- rnorm(100)
        hist(data, main = "Histogram of Data", xlab = "Values", col = "blue", border = "black")
```

### Box Plot

```{r, echo=FALSE, message=FALSE}
 # Create a box plot in R
        data <- rnorm(100)
        boxplot(data, main = "Box Plot Example")
```

-   **Definition:** Graphically represents data distribution based on quartiles, highlighting outliers, median, and spread.
-   **Pros:**
    -   Effective for identifying outliers.
    -   Compares multiple groups easily.
    -   Robust to non-normal data.
-   **Cons:**
    -   Less informative for small data sets.
    -   Not ideal for displaying multimodal distributions.

```{r, eval=FALSE}
# Create a box plot in R
        data <- rnorm(100)
        boxplot(data, main = "Box Plot Example")
```

### Pie Chart

```{r, echo=FALSE, message=FALSE}
 # Create a pie chart in R
        values <- c(10, 20, 30, 40)
        labels <- c("A", "B", "C", "D")
        pie(values, labels = labels, main = "Pie Chart Example")
```

-   **Definition:** Displays relative proportions in part-to-whole relationships using slices of a circle.
-   **Pros:**
    -   Visually intuitive.
    -   Effective for simple categorical data.
-   **Cons:**
    -   Difficult to compare proportions across different charts.
    -   Not suitable for large numbers of categories.
    -   Can distort differences due to angle perception issues.

```{r, eval=FALSE}
# Create a pie chart in R
        values <- c(10, 20, 30, 40)
        labels <- c("A", "B", "C", "D")
        pie(values, labels = labels, main = "Pie Chart Example")
```

### Scatter Plot

```{r, echo=FALSE, message=FALSE}
# Create a scatter plot in R
        x <- rnorm(100)
        y <- rnorm(100)
        plot(x, y, main = "Scatter Plot Example", xlab = "X-axis", ylab = "Y-axis")
```

-   **Definition:** Shows the relationship between two continuous variables.
-   **Pros:**
    -   Identifies correlations and relationships between variables.
    -   Highlights clusters and patterns.
    -   Detects outliers effectively.
-   **Cons:**
    -   Can suffer from overplotting with large data sets.
    -   Requires understanding of correlation interpretation.

```{r, eval=FALSE}
# Create a scatter plot in R
        x <- rnorm(100)
        y <- rnorm(100)
        plot(x, y, main = "Scatter Plot Example", xlab = "X-axis", ylab = "Y-axis")
```

### Mosaic Plot

```{r, echo=FALSE, message=FALSE}
# Load the Titanic dataset
data(Titanic)

mosaicplot(~ Sex +Survived, data = Titanic, color = TRUE)

```

-   **Definition:** Displays the relationships between two or more categorical variables using a stacked rectangle visualization.
-   **Pros:**
    -   Effective for identifying associations in contingency tables.
    -   Highlights interactions between variables.
-   **Cons:**
    -   Interpretation can be challenging for complex relationships.
    -   Less intuitive than simpler visualizations.

```{r, eval=FALSE}
# Create a mosaic plot in R
        library(vcd)
        data(Titanic)
        mosaic(Titanic, shade = TRUE, legend = TRUE)
```

### Bar/Column Chart

```{r, echo=FALSE, message=FALSE}
# Create a bar chart in R
        categories <- c("A", "B", "C", "D")
        values <- c(10, 20, 30, 40)
        barplot(values, names.arg = categories, main = "Bar Chart Example", col = "blue")
```

-   **Definition:** A bar or column chart compares data across categories using rectangular bars, where each bar's length represents the value or frequency of a category. The bars can be vertical (column chart) or horizontal (bar chart).

-   **Pros:**

    -   **Easily Compares Data:**
        -   Makes it simple to compare data across different categories.
        -   Highlights significant differences between groups.
    -   **Quick Insight:**
        -   Provides immediate visual insights into category differences.
        -   Suitable for presenting data to non-technical audiences.
    -   **Intuitive Interpretation:**
        -   Clear and straightforward representation.
        -   Suitable for a wide range of audiences due to its intuitive nature.
    -   **Flexibility:**
        -   Can represent both frequency (count data) and summary statistics (e.g., means, medians).
        -   Can be customized with different bar colors, stacking, grouping, etc.

-   **Cons:**

    -   **Misleading Scaling:**
        -   Can mislead if bar lengths, axis scales, or data representations are inconsistent.
        -   Improper axis truncation may exaggerate or minimize differences.
    -   **Overloading with Categories:**
        -   Too many categories can clutter the chart, making it hard to interpret.
        -   Limited space can lead to overlapping labels, hindering readability.
    -   **Not Ideal for Continuous Data:**
        -   Works best with categorical data but not suitable for continuous variables.
        -   Summarizing continuous data into categories can lead to loss of detail or oversimplification.
    -   **Chart Junk:**
        -   Excessive use of gridlines, 3D effects, or non-data elements can lead to "chart junk," distracting from the data itself.
    -   **Data-to-Ink Ratio:**
        -   Low data-to-ink ratio due to large bars relative to the actual information conveyed.

**Summary:**\
Bar/column charts are highly effective for comparing categorical data and highlighting category differences. However, careful attention to scaling and data representation is crucial to ensure accurate and clear visualization.

```{r, eval=FALSE}
# Create a bar chart in R
        categories <- c("A", "B", "C", "D")
        values <- c(10, 20, 30, 40)
        barplot(values, names.arg = categories, main = "Bar Chart Example", col = "blue")
```

### Quantile Plot

```{r, echo=FALSE, message=FALSE}
 # Create a quantile plot in R
        data <- rnorm(100)
        qqnorm(data)
        qqline(data, col = "blue")
```

-   **Definition:** Displays cumulative quantiles of a distribution versus expected quantiles (e.g., normal distribution).

-   **Pros:**

    -   **Identifies Distribution Deviations:**
        -   Clearly shows deviations of data distribution from a theoretical distribution.
    -   **Efficient Distribution Fit Determination:**
        -   Quickly determines whether the data fits a specific distribution.

-   **Cons:**

    -   **Requires Statistical Knowledge:**
        -   Accurate interpretation requires understanding quantile statistics.
    -   **Less Intuitive:**
        -   Less intuitive than histograms or box plots for non-technical audiences.

**Summary:**\
Quantile plots are ideal for identifying distribution deviations from theoretical distributions. However, they require statistical knowledge for accurate interpretation and might be less intuitive than other visualization techniques.

```{r, eval=FALSE}
# Create a quantile plot in R
        data <- rnorm(100)
        qqnorm(data)
        qqline(data, col = "blue")
```

### Treemap

```{r, echo=FALSE, message=FALSE}
# Create a treemap in R
        library(treemap)
        
        data <- data.frame(
          category = c("A", "B", "C", "D"),
          subcategory = c("A1", "B1", "C1", "D1"),
          value = c(10, 20, 30, 40)
        )
        
        treemap(
          data,
          index = c("category", "subcategory"),
          vSize = "value",
          title = "Treemap Example"
        )
```

-   **Definition:** Visualizes hierarchical data using nested rectangles.

-   **Pros:**

    -   **Displays Large Amounts of Data Efficiently:**
        -   Suitable for representing large hierarchical data sets.
    -   **Intuitive for Showing Hierarchical Relationships:**
        -   Clearly shows hierarchical relationships in a visual format.
    -   **Highlights Patterns and Clusters Visually:**
        -   Highlights patterns, clusters, and outliers effectively.

-   **Cons:**

    -   **Distorts Proportions if Sizes are Too Small:**
        -   Smaller rectangles can become unreadable and distort proportions.
    -   **Challenging to Interpret for High-Depth Hierarchies:**
        -   Deep hierarchies can make interpretation challenging and confusing.

**Summary:**\
Treemaps are excellent for visualizing hierarchical data efficiently and intuitively. However, careful design is crucial to avoid distortions and ensure interpretability, especially with deep hierarchies.

```{r, eval=FALSE}
# Create a treemap in R
        library(treemap)
        
        data <- data.frame(
          category = c("A", "B", "C", "D"),
          subcategory = c("A1", "B1", "C1", "D1"),
          value = c(10, 20, 30, 40)
        )
        
        treemap(
          data,
          index = c("category", "subcategory"),
          vSize = "value",
          title = "Treemap Example"
        )
```

## Advanced Graphs

### Time Series Plots

```{r, echo=FALSE, message=FALSE}
       # Create a time series plot in R
        time <- seq(from = as.Date("2022-01-01"), by = "month", length.out = 12)
        values <- rnorm(12, mean = 50, sd = 10)
        plot(time, values, type = "l", main = "Time Series Plot Example")
```

-   **Definition:** Displays data trends over time.
-   **Pros:**
    -   Ideal for identifying trends, cycles, and seasonal patterns.
    -   Supports multiple time series on the same graph.
    -   Facilitates forecasting and future planning.
-   **Cons:**
    -   Less effective for non-temporal data.
    -   Can suffer from overplotting with too many series.

```{r, eval=FALSE}
# Create a time series plot in R
        time <- seq(from = as.Date("2022-01-01"), by = "month", length.out = 12)
        values <- rnorm(12, mean = 50, sd = 10)
        plot(time, values, type = "l", main = "Time Series Plot Example")
```

### Maps

```{r, echo=FALSE, message=FALSE}
# Create a map in R
        library(ggplot2)
        library(maps)

        world_map <- map_data("world")
        ggplot(world_map, aes(x = long, y = lat, group = group)) +
          geom_polygon(fill = "lightblue", color = "white") +
          ggtitle("World Map Example")
```

-   **Definition:** Visualizes geographic data by plotting locations on maps.
-   **Pros:**
    -   Provides spatial context to data.
    -   Effective for highlighting geographic patterns.
    -   Supports data aggregation at regional levels.
-   **Cons:**
    -   Interpretation can be misleading without proper scaling.
    -   Requires geographic knowledge for accurate interpretation.

```{r, eval=FALSE}
# Create a map in R
        library(ggplot2)
        library(maps)

        world_map <- map_data("world")
        ggplot(world_map, aes(x = long, y = lat, group = group)) +
          geom_polygon(fill = "lightblue", color = "white") +
          ggtitle("World Map Example")
```

### Unstructured Text

#### Word Cloud Example

<!-- ```{r, echo=FALSE, message=FALSE} -->

<!-- # Summarize unstructured text in R -->

<!--         library(tm) -->

<!--         library(wordcloud) -->

<!--         text <- c("sepsis", "infection", "fever", "hospital", "antibiotics",  -->

<!--           "bacteria", "blood", "severe", "septic shock", "intensive care") -->

<!--         # Repeat each word a random number of times -->

<!--         text_repeated <- unlist(lapply(text, function(word) rep(word, sample(5:15, 1)))) -->

<!--         corpus <- Corpus(VectorSource(text_repeated)) -->

<!--         tdm <- TermDocumentMatrix(corpus) -->

<!--         word_freq <- sort(rowSums(as.matrix(tdm)), decreasing = TRUE) -->

<!--         wordcloud(names(word_freq), word_freq, scale = c(3, 0.5), colors = brewer.pal(8, "Dark2")) -->

<!-- ``` -->

-   **Definition:** Summarizes text data numerically to gain insights.
-   **Pros:**
    -   Extracts useful information from textual data.
    -   Identifies patterns and trends in large text corpora.
    -   Supports topic modeling and sentiment analysis.
-   **Cons:**
    -   Requires preprocessing to handle noise and inconsistencies.
    -   Interpretation challenges due to language ambiguities.

```{r, eval=FALSE}
# Summarize unstructured text in R
        library(tm)
        library(wordcloud)

        text <- c("word1", "word2", "word3", "word1", "word2")
        corpus <- Corpus(VectorSource(text))
        tdm <- TermDocumentMatrix(corpus)
        word_freq <- sort(rowSums(as.matrix(tdm)), decreasing = TRUE)
        wordcloud(names(word_freq), word_freq, scale = c(3, 0.5), colors = brewer.pal(8, "Dark2"))
```

### References

1.  Tufte, Edward R. *The Visual Display of Quantitative Information*. Graphics Press, 2001.
2.  Few, Stephen. *Now You See It: Simple Visualization Techniques for Quantitative Analysis*. Analytics Press, 2009.
