---
title: "Introduction to Simple Linear Regression"
author: "Dr. Joshua Lambert"
format:
  html:
    toc: true
    toc-title: "Table of Contents"
    code-fold: false
    code-tools: true
    css: ../styles.css
editor: 
  markdown: 
    wrap: 72
---

# Introduction to Simple Linear Regression

## Objectives

-   Understand the fundamentals of Simple Linear Regression
-   Learn the definition and purpose of Simple Linear Regression
-   Identify key concepts and assumptions of Simple Linear Regression
-   Visualize the relationship between variables
-   Calculate and interpret correlation coefficients
-   Formulate the regression equation
-   Estimate the regression coefficients
-   Test the significance of the regression coefficients
-   Calculate confidence intervals for the coefficients
-   Understand the coefficient of determination (R-squared)
-   Perform residual analysis and diagnostic plots
-   Interpret regression coefficients
-   Check assumptions
-   Explore examples of Simple Linear Regression in nursing research

## Definition and Purpose of Simple Linear Regression

Simple Linear Regression is a statistical method used to model the
relationship between a single independent variable (predictor) and a
continuous dependent variable (outcome). It aims to predict the
dependent variable based on the values of the independent variable.

### Key Concepts and Assumptions of Simple Linear Regression

1.  **Linearity:** The relationship between the predictor and outcome is
    linear.
2.  **Independence:** Observations are independent.
3.  **Homoscedasticity:** The variance of residuals is constant.
4.  **Normality:** Residuals are normally distributed.

## Visualizing the Relationship Between Variables

Use scatter plots to visualize the relationship between the predictor
and outcome variables. For more on creating these visualizations in R, see the [Graphing Your Data](graphing-your-data.qmd) lecture.

### Example Visualization in R

```{r, eval=FALSE}
# Scatter plot in R
set.seed(123)
x <- rnorm(100)
y <- 3 * x + rnorm(100)

plot(x, y, main = "Scatter Plot of Predictor vs. Outcome", xlab = "Predictor (x)", ylab = "Outcome (y)", col = "blue")
```

## Calculating and Interpreting Correlation Coefficients

### Pearson Correlation Coefficient (( r ))

Measures the strength and direction of the linear relationship between
two continuous variables.

### Formula

$$
r = \frac{\sum (x_i - \bar{x})(y_i - \bar{y})}{\sqrt{\sum (x_i - \bar{x})^2 \sum (y_i - \bar{y})^2}}
$$

### Interpretation

-   **( r = +1 ):** Perfect positive correlation
-   **( r = -1 ):** Perfect negative correlation
-   **( r = 0 ):** No correlation

## Formulating the Regression Equation

The regression equation represents the linear relationship between the
predictor (( x )) and outcome (( y )).

### Equation

$$
y = \beta_0 + \beta_1 x
$$ where: - ( \beta\_0 ): Intercept (constant) - ( \beta\_1 ): Slope
(coefficient)

## Estimating the Regression Coefficients

The coefficients are estimated using the method of least squares, which
minimizes the sum of squared residuals.

### Formula

-   **Slope (Coefficient):** $$
    \beta_1 = \frac{\sum (x_i - \bar{x})(y_i - \bar{y})}{\sum (x_i - \bar{x})^2}
    $$

-   **Intercept (Constant):** $$
    \beta_0 = \bar{y} - \beta_1 \bar{x}
    $$

## Testing the Significance of the Regression Coefficients

### Hypothesis Testing

-   **Null Hypothesis (( H_0 )):** The coefficient is not significantly
    different from zero (( \beta\_1 = 0 )).
-   **Alternative Hypothesis (( H_1 )):** The coefficient is
    significantly different from zero (( \beta\_1 \neq 0 )).

### t-Test

The significance of the coefficients is tested using a t-test.

### Formula

$$
t = \frac{\hat{\beta}}{SE(\hat{\beta})}
$$ where: - ( \hat{\beta} ): Estimated coefficient - ( SE(\hat{\beta})
): Standard error of the coefficient

## Confidence Intervals for the Coefficients

A confidence interval provides a range within which the true coefficient
is likely to fall.

### Formula

$$
\hat{\beta} \pm t^* \cdot SE(\hat{\beta})
$$ where: - ( t\^\* ): Critical value from the t-distribution

## Coefficient of Determination (R-squared)

Measures the proportion of the variance in the outcome variable
explained by the predictor.

### Formula

$$
R^2 = \frac{\sum (\hat{y} - \bar{y})^2}{\sum (y_i - \bar{y})^2}
$$

## Residual Analysis and Diagnostic Plots

### Residual Plots

Plot residuals against the predicted values to check for
homoscedasticity and non-linearity.

### Q-Q Plot

Check the normality of residuals by plotting them against a normal
distribution.

### Example in R

```{r, eval=FALSE}
# Residual plots and Q-Q plot in R
fit <- lm(y ~ x)

# Residual vs. Fitted plot
plot(fit, which = 1, main = "Residuals vs. Fitted")

# Q-Q plot
plot(fit, which = 2, main = "Q-Q Plot")
```

## Interpretation of Regression Coefficients

-   \*\*Intercept (( \beta\_0 )):\*\* The expected value of the outcome
    when the predictor is zero.
-   \*\*Slope (( \beta\_1 )):\*\* The expected change in the outcome for
    a one-unit increase in the predictor.

## Checking Assumptions

-   **Linearity:** Residual vs. Fitted plot should show a random
    pattern.
-   **Independence:** Observations should be independent.
-   **Homoscedasticity:** Residual vs. Fitted plot should show constant
    variance.
-   **Normality:** Q-Q plot should show residuals roughly following a
    straight line.

## Examples of Simple Linear Regression in Nursing Research

### Example 1: Evaluating the Relationship Between BMI and Blood Pressure

-   **Research Question:** How does BMI affect blood pressure in adults?
-   **Predictor (x):** BMI
-   **Outcome (y):** Blood Pressure

<a href="slr_ex1.csv" download>download example data (CSV)</a>

```{r, eval=FALSE}
# Example in R
# Simulated data
set.seed(123)
BMI <- rnorm(100, mean = 25, sd = 3)
BP <- 2 * BMI + rnorm(100, sd = 5)

# Simple Linear Regression
fit1 <- lm(BP ~ BMI)
summary(fit1)

# Plot
plot(BMI, BP, main = "BMI vs. Blood Pressure", xlab = "BMI", ylab = "Blood Pressure")
abline(fit1, col = "red")
```

```{=html}
<script>
function revealAnswer(card) {
  var question = card.querySelector('.question');
  var answer = card.querySelector('.answer');
  question.style.display = 'none';
  answer.style.display = 'block';
}
</script>
```
