---
title: "Introduction to Two-Way ANOVA"
author: "Dr. Joshua Lambert"
format:
  html:
    toc: true
    toc-title: "Table of Contents"
    code-fold: false
    code-tools: true
    css: ../styles.css
editor: 
  markdown: 
    wrap: 72
---

[**Download R Code for this Lecture**](code/introduction-to-two-way-anova.R){ .btn .btn-primary }



# Introduction to Two-Way ANOVA

## Objectives

-   Understand the fundamentals of Two-Way ANOVA
-   Learn the definition and purpose of Two-Way ANOVA
-   Identify the key differences between One-Way ANOVA and Two-Way ANOVA
-   Understand the structure of a Two-Way ANOVA design
-   Recognize main effects and interaction
-   Comprehend the assumptions of Two-Way ANOVA
-   Calculate Sum of Squares (SS) and Mean Squares (MS)
-   Conduct hypothesis testing for main effects and interaction
-   Measure effect sizes and significance
-   Interpret main effects and interaction
-   Conduct post-hoc analysis
-   Identify outliers and influential observations, normality, and
    homogeneity
-   Explore examples of Two-Way ANOVA in nursing research

## Definition and Purpose of Two-Way ANOVA

Two-Way ANOVA is a statistical test used to examine the impact of two
independent variables on a continuous dependent variable. It is an
extension of One-Way ANOVA but allows for the simultaneous evaluation of
two factors, making it suitable for analyzing more complex experimental
designs.

### Key Differences Between One-Way and Two-Way ANOVA

-   **One-Way ANOVA:** Evaluates the effect of a single independent
    variable on a continuous outcome.
-   **Two-Way ANOVA:** Evaluates the effects of two independent
    variables (factors), including any interaction between them.

## Understanding the Structure of a Two-Way ANOVA Design

In a Two-Way ANOVA, each factor (independent variable) can have multiple
levels. For example, in a clinical trial examining the effects of drug
dosage (Factor 1) and diet (Factor 2), the design might include: -
**Factor 1 (Dosage):** Low, Medium, High - **Factor 2 (Diet):** Control,
Low-Fat, High-Fat

### Main Effects and Interaction

-   **Main Effect:** The impact of one factor on the dependent variable,
    ignoring the levels of the other factor.
-   **Interaction Effect:** The combined effect of both factors on the
    dependent variable that is different from the sum of their
    individual effects.

## Assumptions of Two-Way ANOVA

### Normality of Residuals

Residuals (errors) should be normally distributed.

### Homogeneity of Variances

Variances should be approximately equal across all levels of the
factors.

## Calculation of SS (Sum of Squares) and MS (Mean Squares)

### Sum of Squares

-   **Total Sum of Squares (SST):** $$
    SST = \sum_{i=1}^{n} (Y_i - \bar{Y})^2
    $$ where ( Y_i ) is each individual observation, and ( \bar{Y} ) is
    the overall mean.

-   **Sum of Squares for Factor A (SSA):** $$
    SSA = \sum_{j=1}^{a} n_{j} (\bar{Y}_{j\cdot} - \bar{Y})^2
    $$ where ( n\_{j} ) is the number of observations in each level of
    Factor A, ( \bar{Y}\_{j\cdot} ) is the mean for each level of Factor

    A.  

-   **Sum of Squares for Factor B (SSB):** $$
    SSB = \sum_{k=1}^{b} n_{k} (\bar{Y}_{\cdot k} - \bar{Y})^2
    $$ where ( n\_{k} ) is the number of observations in each level of
    Factor B, ( \bar{Y}\_{\cdot k} ) is the mean for each level of
    Factor B.

-   **Sum of Squares for Interaction (SSAB):** $$
    SSAB = \sum_{j=1}^{a} \sum_{k=1}^{b} n_{jk} (\bar{Y}_{jk} - \bar{Y}_{j\cdot} - \bar{Y}_{\cdot k} + \bar{Y})^2
    $$ where ( \bar{Y}\_{jk} ) is the mean of each combination of levels
    in Factors A and B.

-   **Sum of Squares for Error (SSE):** $$
    SSE = \sum_{j=1}^{a} \sum_{k=1}^{b} \sum_{i=1}^{n_{jk}} (Y_{ijk} - \bar{Y}_{jk})^2
    $$

### Mean Squares

Mean Squares are obtained by dividing each Sum of Squares by the
corresponding degrees of freedom.

-   **Mean Square for Factor A (MSA):** $$
    MSA = \frac{SSA}{a - 1}
    $$

-   **Mean Square for Factor B (MSB):** $$
    MSB = \frac{SSB}{b - 1}
    $$

-   **Mean Square for Interaction (MSAB):** $$
    MSAB = \frac{SSAB}{(a - 1)(b - 1)}
    $$

-   **Mean Square for Error (MSE):** $$
    MSE = \frac{SSE}{ab(n - 1)}
    $$

## Hypothesis Testing for Main Effects and Interaction

### Hypotheses

-   **Factor A (Main Effect):**
    -   Null Hypothesis (( H_0 )): All levels of Factor A have the same
        mean.
    -   Alternative Hypothesis (( H_1 )): At least one level of Factor A
        has a different mean.
-   **Factor B (Main Effect):**
    -   Null Hypothesis (( H_0 )): All levels of Factor B have the same
        mean.
    -   Alternative Hypothesis (( H_1 )): At least one level of Factor B
        has a different mean.
-   **Interaction Effect:**
    -   Null Hypothesis (( H_0 )): No interaction between Factors A and
        B.  
    -   Alternative Hypothesis (( H_1 )): Interaction exists between
        Factors A and B.

### F-Ratio Calculations

-   **Factor A:** $$
    F_A = \frac{MSA}{MSE}
    $$

-   **Factor B:** $$
    F_B = \frac{MSB}{MSE}
    $$

-   **Interaction (A x B):** $$
    F_{AB} = \frac{MSAB}{MSE}
    $$

### Interpretation

-   Compare the F-ratios to the critical values from the F-distribution.
-   If the calculated F-ratio is greater than the critical value, reject
    the null hypothesis.

<iframe src="http://modeling.nursing.uc.edu/anova_2w_app_1/" width="100%" height="1000px">

</iframe>

## Effect Sizes and Significance

### Eta Squared (( \eta\^2 ))

Represents the proportion of the total variance explained by each
factor.

-   **Factor A:** $$
    \eta^2_A = \frac{SSA}{SST}
    $$

-   **Factor B:** $$
    \eta^2_B = \frac{SSB}{SST}
    $$

-   **Interaction (A x B):** $$
    \eta^2_{AB} = \frac{SSAB}{SST}
    $$

## Post-Hoc Analysis

If significant main or interaction effects are found, post-hoc tests
(e.g., Tukey's HSD) can identify which specific groups differ.

### Tukey's HSD

$$
HSD = q \sqrt{\frac{MSE}{n}}
$$ where ( q ) is the studentized range statistic, and ( n ) is the
sample size per group.

## Identifying Outliers and Influential Observations

### Normality

-   Use Q-Q plots to check the normality of residuals.

### Homogeneity of Variances

-   Use Levene's test or Bartlett's test.

### Outliers

-   Use residual plots or Cook's distance to identify influential
    points.

## Examples of Two-Way ANOVA in Nursing Research

### Example 1: Evaluating Drug Dosage and Diet Effects

-   **Research Question:** How do drug dosage and diet affect blood
    pressure levels in patients?
-   **Factors:**
    -   Dosage (Low, Medium, High)
    -   Diet (Control, Low-Fat, High-Fat)
-   **Response Variable:** Blood Pressure Levels

```{r, eval=FALSE}
# Example in R
# Simulated data
set.seed(123)
data <- data.frame(
  Dosage = factor(rep(c("Low", "Medium", "High"), each = 30)),
  Diet = factor(rep(c("Control", "Low-Fat", "High-Fat"), times = 30)),
  BP = rnorm(90, mean = rep(c(120, 115, 130), each = 30), sd = 10)
)

# Two-Way ANOVA
model <- aov(BP ~ Dosage * Diet, data = data)
summary(model)
```

### Example 2: Nursing Intervention and Patient Demographics

-   **Research Question:** How do a nursing intervention and patient
    demographics affect recovery time?
-   **Factors:**
    -   Intervention (Control, Intensive)
    -   Gender (Male, Female)
-   **Response Variable:** Recovery Time

```{r, eval=FALSE}
# Example in R
# Simulated data
set.seed(456)
data2 <- data.frame(
  Intervention = factor(rep(c("Control", "Intensive"), each = 50)),
  Gender = factor(rep(c("Male", "Female"), times = 50)),
  RecoveryTime = rnorm(100, mean = rep(c(14, 12), each = 50), sd = 3)
)

# Two-Way ANOVA
model2 <- aov(RecoveryTime ~ Intervention * Gender, data = data2)
summary(model2)
```

