---
title: "NPHD 9040 Homework 3"
subtitle: "Two-Way ANOVA"
format: 
  html:
    toc: true
    css: ../styles.css
---
::: {.callout-tip icon=false}
## New to R Markdown?
[**Click here for a guide on Getting Started with R Markdown**](../Other/working_with_rmd.qmd)
:::

::: {.callout-note icon=false}
## Downloads
[**Download Assignment Template**](_templates/hw3.Rmd){ .btn .btn-primary }

[**Download Data: Coffee Data.csv**](../course_data/Coffee Data.csv){ .btn .btn-success }

:::


# Instructions

1.  **Download** the assignment template (right-click and "Save Link As..." on the button above).
2.  **Open** the `.Rmd` file in RStudio.
3.  **Complete** the assignment by adding your R code chunks and written answers where indicated.
4.  **Submission**: Submit **BOTH** your completed R Markdown file (`.Rmd`) and the rendered output file (HTML, PDF, or Word) to Canvas.

**Goal**: This assignment covers Two-Way ANOVA, allowing you to examine the main effects of two independent variables and their interaction on a continuous outcome.

# Question 1: Data Import and Exploration

**Task**: Prepare the data.

1.  Load `Coffee Data.csv`.
2.  Ensure `Grind` and `Temperature` are treated as factors.

**Produce**:
*   **Code**: Loading and conversion code.
*   **Output**: `str()` or `summary()` showing the variables are Factors.

::: {.callout-tip}
## Hint
Remember to use `as.factor()` if variables aren't read correctly. See [Data Types](../Lectures/getting-started-with-r.qmd#data-types).
:::

```{r}
#| label: load-data
#| warning: false

# Your code here
```

# Question 2: Descriptive Statistics

**Task**: Visualize the interaction.

1.  Calculate the Mean `Strength` for all combinations of Grind and Temperature (e.g., Fine/Hot, Fine/Cold, etc.).
2.  Create an Interaction Plot (line plot with one variable on X and separate lines for the other).

**Produce**:
*   **Code**: Commands for means (e.g., `tapply` or `group_by`) and plotting (e.g., `interaction.plot` or `ggplot`).
*   **Output**: The table of means and the plot.
*   **Interpretation**: Looking at the plot, do the lines cross? Does the effect of Temperature look different depending on Grind?

::: {.callout-tip}
## Hint
See [Example 1 in Two-Way ANOVA](../Lectures/introduction-to-two-way-anova.qmd#examples-of-two-way-anova-in-nursing-research) for relevant code structures.
:::

```{r}
#| label: interaction-plot

# Your code here
```

# Question 3: Two-Way ANOVA Model

**Task**: Test for significance.

1.  Fit the Two-Way ANOVA model: `Strength ~ Grind * Temperature`.
2.  Produce the ANOVA table.

**Produce**:
*   **Code**: `aov()` command and `summary()`.
*   **Output**: The ANOVA table showing Main Effects and Interaction.
*   **Interpretation**:
    *   Is the Interaction effect significant?
    *   Is the Main Effect of Grind significant?
    *   Is the Main Effect of Temperature significant?

::: {.callout-tip}
## Hint
Follow the syntax in [Examples of Two-Way ANOVA](../Lectures/introduction-to-two-way-anova.qmd#examples-of-two-way-anova-in-nursing-research). Note the use of `*` for interaction.
:::

```{r}
#| label: run-2way-anova

# Your code here
```

# Question 4: Post-Hoc Tests (Simple Main Effects)

**Task**: Deconstruct the interaction (or main effects).

1.  If the interaction was significant, finding *which* temperature differs at *which* grind level is important. Run a Tukey HSD test on the model.
2.  (Optional challenge) Attempt to test "Simple Main Effects" (e.g., does Temperature matter specifically for "Fine" grind?).

**Produce**:
*   **Code**: TukeyHSD command.
*   **Output**: Pairwise comparison table.
*   **Interpretation**: Select one significant comparison from the list and explain it in plain English (e.g., "Strength is significantly higher for Hot vs Cold, but only when Grind is Fine").

::: {.callout-tip}
## Hint
For interactions, [Post-hoc Analysis](../Lectures/introduction-to-two-way-anova.qmd#post-hoc-analysis) explains how to interpret the complex pairwise comparisons.
:::

```{r}
#| label: post-hoc-2way

# Your code here
```
