---
title: "NPHD 9040 Homework 2"
subtitle: "One-Way ANOVA"
author: "Your Name Here"
date: "Due Date"
format: 
  html:
    toc: true
---

# Instructions

1.  **Download** the assignment template (right-click and "Save Link As..." on the button above).
2.  **Open** the `.Rmd` file in RStudio.
3.  **Complete** the assignment by adding your R code chunks and written answers where indicated.
4.  **Submission**: Submit **BOTH** your completed R Markdown file (`.Rmd`) and the rendered output file (HTML, PDF, or Word) to Canvas.

**Goal**: This assignment covers the One-Way Analysis of Variance (ANOVA), including assumptions, execution in R, and interpretation of results.

# Question 1: Data Import and Exploration

**Task**: Prepare the data for analysis.

1.  Load `anova_ex1.csv` into R.
2.  Display the first few rows.
3.  Check the structure of the dataframe.

**Produce**:
*   **Code**: R commands to load and inspect data.
*   **Output**: The `head()` and `str()` (or `summary()`) of the dataset.
*   **Interpretation**: Briefly confirm that your grouping variable is a factor (or character) and your outcome is numeric.

::: {.callout-tip}
## Hint
Review [Reading Data](../Lectures/getting-started-with-r.qmd#reading-data-from-csv-files) if needed.
:::

```{r}
#| label: load-data
#| warning: false

# Your code here
```

# Question 2: Descriptive Statistics and Visualization

**Task**: Explore the data visually and numerically before testing.

1.  Calculate determining summary statistics (Mean and SD) for each group.
2.  Create a boxplot of `stress` by `group`.

**Produce**:
*   **Code**: R commands for summary stats and plotting.
*   **Output**: A table (or printed output) of Means/SDs and the Boxplot image.
*   **Interpretation**: Describe any apparent differences in the boxplots (e.g., "The Behavioral group appears to have higher stress...").

::: {.callout-tip}
## Hint
See [Visualizing Using Boxplots](../Lectures/introduction-to-anova.qmd#visualizing-using-boxplots) for code examples.
:::

```{r}
#| label: descriptive-stats

# Your code here
```

# Question 3: Assumptions of ANOVA

**Task**: Verify that the data meets the requirements for ANOVA.

1.  **Normality**: Test the normality of the residuals (Shapiro-Wilk) or inspect a Q-Q plot.
2.  **Homogeneity of Variance**: Test if groups have equal variances (Levene's Test).

**Produce**:
*   **Code**: R commands for Shapiro-Wilk and Levene's tests.
*   **Output**: Test statistics and p-values.
*   **Interpretation**:
    *   Normality: Is the p-value > 0.05? What does that mean?
    *   Homogeneity: Is the p-value > 0.05? What does that mean?

::: {.callout-tip}
## Hint
Check [Assumptions of ANOVA](../Lectures/introduction-to-anova.qmd#assumptions-of-anova) and how to test them in R.
:::

```{r}
#| label: check-assumptions

# Your code here
```

# Question 4: One-Way ANOVA

**Task**: Perform the omnibus One-Way ANOVA test.

1.  Run the ANOVA model (`aov(outcome ~ group)`).
2.  Print the summary table.

**Produce**:
*   **Code**: R command to fit model and show summary.
*   **Output**: The ANOVA table.
*   **Interpretation**: Report the F-statistic, degrees of freedom, and p-value. State clearly whether the result is statistically significant and what that implies (ie. "At least one group differs...").

::: {.callout-tip}
## Hint
Follow the steps in [Conducting ANOVA in R](../Lectures/introduction-to-anova.qmd#conducting-anova-in-r).
:::

```{r}
#| label: run-anova

# Your code here
```

# Question 5: Post-Hoc Analysis

**Task**: Identifying specific group differences.

1.  If the ANOVA in Q4 was significant, conduct a Tukey HSD post-hoc test.
2.  If it was not significant, note that post-hoc tests are not needed (but run the code anyway for practice).

**Produce**:
*   **Code**: R command for TukeyHSD.
*   **Output**: The table of pairwise comparisons.
*   **Interpretation**: Identify which specific pairs of groups are significantly different from each other based on the adjusted p-values.

::: {.callout-tip}
## Hint
See [Post-hoc Analysis](../Lectures/introduction-to-anova.qmd#post-hoc-analysis) for using `TukeyHSD()`.
:::

```{r}
#| label: post-hoc

# Your code here
```
